Predictive modeling using genomic annotations

### 2018-09-12 Manuscript/manuscript.Rmd
updates

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect several epigenetic functions (Aiden et al.). Analyses have shown that the genome is tightly compacted into distinct compartments.  There exist regions within these compartments that that have been shown to be highly conserved and self-interacting, and termed topologically associating domains (TADs) (Dixon et al.). Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD (Symmons et al.). This suggests that the boundaries of TADs may play a role in restricting the function of elements such as enhancers, thereby impacting the transcription of genes.

More recently, it has been discovered that insulator sequences have a primary role in orchestrating the topological arrangement of higher-order chromatin architecture (Phillips-Cremins et al). Insulators are multi-faceted regulatory sequences that moderate a variety of genomic processes including activation, repression, and enhancer blocking. Specifically, the insulator binding protein CTCF has been found to be enriched at the boundary sequences of topologically associating domains in human cells and may therefore act as a mediator of long range chromatin contacts (Zuin et al.). Furthermore, other regulatory elements like DNase I hypersensitive sites, the H3K36 trimethylation, and transcription factors such as ZNF274 and YY1 have been more frequently observed at TAD boundaries than in other regions (Hong et al). Each of these associated with more open chromatin which alter the accessibility of genes for transcription.   [??? They identify more; !!! included mention of histone mods. and TFs]. The distinct patterns of some of these different proteins and functional elements point toward the opportunity of computational approaches in predicting the development of TAD boundaries.

Due to the size of Hi-C data and the abundance of available genomic features, few methods have been developed to study the role and interplay of different sets of these features on the folding of DNA. Furthermore, many accepted methods ignore key characteristics of the data that, when addressed, may improve predictive performances. One such pitfall is the issue of highly imbalanced classes. The commonality between most accepted methods for predicting TAD boundaries considers binning the genome and labeling the bins as either containing a TAD boundary or not. However, due to the sparcity of TADs throughout the genome, most bins will not contain a TAD boundary. This will contribute to highly imbalanced classes and will also introduce predictors with near-zero variance. It has been shown that imbalanced classes can result in poor predictive performances of classification algorithms and unless the class imbalance is mild, researchers should always use an appropriate method for dealing with this (Blagus et al). Another problem lies in identifying which positional characteristics of the genomic features will optimize prediction. It is unclear if distances or overlaps between genomic bins and regions of functional elements contain the most information. Lastly, given that some features tend to colocalize near TAD boundaries, most predictive models may be unable to identify the complete set of features most associated with boundary formation, and it is important that the proposed model is able to handle multiple correlated features.

In a previous study (Mourad et al) both a multiple logistic regression model and a LASSO regularized regression model were applied to a wide array of architectural proteins from the GM12878 cell line to predict TAD boundaries. The set of features consisted of whether or not the coordinates of a functional element overlaped with a particular genomic bin (yes or no). In the case of a partial overlap, a percentage of overlap was used. However, the class imbalance issue was not addressed. Additionally, neither model is suited to handle multiple correlated predictors. Furthermore, ROC curves and AUCs were used to evaluate model performance. This is problematic, especially for highly imbalaned classes because it dismisses how well the model is able to identify true positive cases. In another study (Huang et al), a more robust model in the form of a Bayesian Additive Regression Trees (BART) model was used. The set of features consisted of a score that averaged the number of sequence reads for each functional element over a particular bin for the GM12878 cell line. Here, the balanced classes were manually created by dividing the positive and negative classes into two equal subsets used for model training and testing. However, it is unclear what balancing technique was used. In addition to this, the model was only performed on a small set of histone marks while ignoring wider classes of epigenomic elements. Lastly, in a more recent study (Hong et al), a bayesian rigde regression model was iteratively used on a set of transcription factor binding sites, transcription start sites, and DNase I hypersensitive sites from the hESC cell line until the best combination of features led to the optimal prediction of TAD boundaries.  Here, the features of the model consisted of the number of occurences of each genomic element within each genomic bin. Balanced classes were manually created by randomly sampling from the set of bins that did not contain a TAD boundary. Given the structure of the search algorithm proposed, though, it can be computationally inefficient to have to iterate through a large set of genomic features to find the particular set that contributes to the best prediction. Because of that, only a maximum set of 10 genomic features were considered at any given time.

The types of models and the feature characteristics used in the aforementioned methods vary considerably. Additionally, a notable omission was the use of distances between regions of the functional elements and genomic bins. Because of the different cell lines available for analysis and the widening list of functional genomic elements, it can be quite useful to have a unified approach in predicting the formation of TAD boundaries. Therefore, in addition to proposing a novel data pre-processing pipeline that addresses irregularities in the data, we also apply an elastic-net model (and/or random forest?) using both distances and counts of overlaps between bins and features as predictors in our model. We aim to assess both the predictive performance of the model and the influence of genomic features on the borders that demarcate topological domains.[??? Clarity. Define 1) the problem, and 2) ways to solve it, which is modeling. Then, for modeling, describe problems connecting genomics+machine learning terms]

>>>>

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect several epigenetic functions (Aiden et al.). Analyses have shown that the genome is tightly compacted into distinct compartments.  There exist regions within these compartments that have been shown to be highly conserved and self-interacting, and termed topologically associating domains (TADs) (Dixon et al.). Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD (Symmons et al.). This suggests that the boundaries of TADs may play a role in restricting the function of elements such as enhancers, thereby impacting the transcription of genes. Furthermore, changing the 3D structure of the DNA and distrupting these domains can lead to adverse outcomes and diseases like cancer. Therefore it is important to be able to identify the key molecular drivers of the formation of TADs in order to further our understanding of the human genome.

The mechanisms underlying the formation of TADs are a complex and active area of research. Recently, it has been discovered that insulator sequences have a primary role in orchestrating the topological arrangement of higher-order chromatin architecture (Phillips-Cremins et al). Insulators are multi-faceted regulatory sequences that moderate a variety of genomic processes including activation, repression, and enhancer blocking. Specifically, the insulator binding protein CTCF has been found to be enriched at the boundary sequences of topologically associating domains in human cells and may therefore act as a mediator of long range chromatin contacts (Zuin et al.). Other regulatory elements like DNase I hypersensitive sites, the H3K36 trimethylation, and transcription factors such as ZNF274 and YY1 have been more frequently observed at TAD boundaries than in other regions (Hong et al). These have been found to be associated with more open chromatin which alter the accessibility of genes for transcription. The distinct patterns of some of these different proteins and functional elements point toward the opportunity of computational approaches in predicting the development of TAD boundaries, which, in genereal, will allow us to better understand what leads to their formation.

Due to the size of Hi-C data and the abundance of available genomic features, few methods have been developed to study the role and interplay of different sets of these features on the folding of DNA. Furthermore, many accepted methods ignore key characteristics of the data that, when addressed, may improve predictive performances. One such pitfall is the issue of highly imbalanced classes. The commonality between most accepted methods for predicting TAD boundaries considers binning the genome and labeling the bins as either containing a TAD boundary or not. However, due to the sparcity of TADs throughout the genome, most bins will not contain a TAD boundary. This will contribute to highly imbalanced classes and will also introduce predictors with near-zero variance. It has been shown that imbalanced classes can result in poor predictive performances of classification algorithms and unless the class imbalance is mild, researchers should always use an appropriate method for dealing with this (Blagus et al).  Furthermore, the inclusion of predictors with near-zero variance can add unnecessary noise to the model and decrease computational efficiency of predictive algorithms. Another problem lies in identifying which positional characteristics of the genomic features will optimize prediction. It is unclear if distances or overlaps between genomic bins and regions of functional elements contain the most information and will lead to better predictive performances. Lastly, given that some features tend to colocalize near TAD boundaries, most predictive models may be unable to identify the complete set of features most associated with boundary formation. This can lead to less informative results and restrict our knowledge of how these genomic elements correlate with eachother.

In a previous study (Mourad et al) both a multiple logistic regression model and a LASSO regularized regression model were applied to a wide array of architectural proteins from the GM12878 cell line to predict TAD boundaries. The set of features consisted of whether or not the coordinates of a functional element overlaped with a particular genomic bin (yes or no). In the case of a partial overlap, a percentage of overlap was used. However, the class imbalance issue was not addressed and neither model used is suited to handle multiple correlated predictors. Additionally, ROC curves and AUCs were used to evaluate model performance, which is problematic, especially for highly imbalaned classes (Hanczar et al). A better indicator of predictive performance with classification algorithms is to consider the F1 score and Matthew's Correlation Coefficient (MCC). In another study (Huang et al), a more robust model in the form of a Bayesian Additive Regression Trees (BART) model was used. The set of features consisted of a score that averaged the number of sequence reads for each functional element over a particular bin for the GM12878 cell line. Here, the class proportion was manually created by dividing the positive and negative classes into two equal subsets used for model training and testing. However, it is unclear what balancing technique was used. In addition to this, the model was only performed on a small set of histone marks while ignoring wider classes of epigenomic elements. Lastly, in a more recent study (Hong et al), a bayesian rigde regression model was iteratively used on a set of transcription factor binding sites, transcription start sites, and DNase I hypersensitive sites from the hESC cell line until the best combination of features led to the optimal prediction of TAD boundaries.  Here, the features of the model consisted of the number of occurences of each genomic element within each genomic bin. Balanced classes were manually created by randomly sampling from the set of bins that did not contain a TAD boundary. Given the structure of the search algorithm proposed, though, it can be computationally inefficient to have to iterate through a large set of genomic features to find the particular set that contributes to the best prediction. Because of that, only a maximum set of 10 genomic features were considered at any given time.

The types of models and the feature characteristics used in the aforementioned methods vary considerably. Because of the different cell lines available for analysis and the widening list of functional genomic elements, it can be quite useful to have a unified approach in predicting the formation of TAD boundaries. Therefore, in this study we have developed a novel data pre-processing pipeline that addresses the irregularities in the TAD boundary data provided by contact matrices at different resolutions. We then performed a variety of machine learning algorithms on the pre-processes data to identify the genomic elements that were most predictive of TAD boundaries. In our models we incorporated the distances from bin to region as features. We found that the models performed on our pre-processed data performed better than previously established methods. Additionally, we were able to confirm that the cis-regulatory elements like insulators and promotors were highly predictive of TAD boundaries. The specific transcriptional repressor, CCCTC-binding factor (CTCF) was found to be highly predictive of TAD boundaries. Other non-transcription factor sites like DNase I hypersensitive sites and certain histone modifications like H3K36me3 were also found to be better predictors.  Our findings confirm that models with pre-processed data and distance-related genomic features not only perform better, but are also able to correctly identify a range of different functional genomic elements associated with the formation of TAD boundaries. This knowlege will allow us to gain better insight into the role that the spatial organisation of the human genome has on gene regulation.
